HO unification from object language to meta language

### 2024-02-21 paper/main.tex
wip intro

\newcommand{\U}{\ensuremath{=_o}\xspace}
\newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

@@

\begin{coqcode}
Instance fin_fin n : Finite (fin n).
Instance nfact_dec n nf : Decision (nfact n nf).
Instance forall_dec A P : Finite A ~$\to$~ 
  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x, P x).
\end{coqcode}

\noindent Under this context the type-class solver is able to prove
the the following statement automatically by back-chaining
the three instances.

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).
\end{coqcode}

\noindent
The encoding of DTT provided by Elpi, that we will discuss at length later in
\cref{sec:encodings,sec:lang-spec}, features the following term constructors:

\begin{elpicode}
kind tm type.
type lam tm -> (tm -> tm) -> tm. % lambda abstraction
type app list tm -> tm.          % n-ary application
type all tm -> (tm -> tm) -> tm. % forall quantifier
type c string -> tm.             % constants
\end{elpicode}

\noindent
\marginpar{TODO: explain HOAS}
Following this term encoding the three instances are represented by the
following rules:

\begin{elpicode}
finite (app[c"fin", N]).
decision (app [c"nfact", N, NF]).
decision (all A x\ app[P, x]) :- finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
\marginpar{TODO: explain pi, cons}
Unfortunately this direct translation of the instances considers the
predicate \coqIn{P} as a first order term. If we try to backchain the
third rule on the encoding of the goal above:

\begin{elpicode}
decision (all (app[c"fin", c"7"]) y\
  app[c"nfact", y, c"3"]).
\end{elpicode}

\noindent
we fail because of this ``higher order'' unification problem (in DTT)
is phrased as a first order unification problem in the meta language.

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y]
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
This time \elpiIn{Pm} is an higher order unification variable (of type
\elpiIn{tm -> tm}). The resulting unification problem is now:

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y
\end{elpicode}

\noindent
That admits one solution:

\begin{elpicode}
Pm = y\ app[c"nfact", y, c"3"]
A = app[c"fin",c"7"]
\end{elpicode}

\noindent
Elpi succeeds in the application of the new rule and then runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment back to the domain of Coq terms (the type \elpiIn{tm}):

@@

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial. Indeed the solution for \elpiIn{P} generates a
(Coq) $\beta$-redex in the second premise (under the \elpiIn{pi x}):

\begin{elpicode}
decision (app[lam A (a\ app[c"nfact", a, c"3"]), x])
\end{elpicode}

\noindent
In turn the redex prevents the second rule to backchain properly since
the following unification problem has no solution:

@@

\noindent
This time the root cause is that the unification procedure of \Ue of the
meta language is not aware of the equational theory of the object logic \U,
even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla..

\section{Alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}

@@

Another alternative

\begin{elpicode}
decision X :- unif X (all A x\ app[P, x]), 
  (pi x\ decision (app[P, x])), finite A.
\end{elpicode}

gives up all half of what the ML gives us. Moreover even if unif here embodies
the eq theory of DTT which is much stronger than the one of the ML, we don't need 
it. According to our experience eta beta suffice, but HO is needed.

>>>>

\newcommand{\U}{\ensuremath{=_o}\xspace}
\newcommand{\nU}{\ensuremath{\neq_o}\xspace}
\newcommand{\Ue}{\ensuremath{=_\lambda}\xspace}
\newcommand{\nUe}{\ensuremath{\neq_\lambda}\xspace}
\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}}
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

@@

\begin{coqcode}
Instance fin_fin n : Finite (fin n).             (* r1 *)
Instance nfact_dec n nf : Decision (nfact n nf). (* r2 *)
Instance forall_dec A P : Finite A ~$\to$~            (* r3 *)
  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
\end{coqcode}

\noindent Under this context of instances a type-class solver is able to prove
the the following statement automatically by back-chaining.

\begin{coqcode}
  Check _ : Decision (forall y: fin 7, nfact y 3).  (* g *)
\end{coqcode}

\noindent
The encoding of DTT provided by Elpi, that we will discuss at length later in
section~\ref{sec:encodings,sec:lang-spec} and ~\ref{}, is an Higher Order Abstract
Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
constructors:

\begin{elpicode}
type lam  tm -> (tm -> tm) -> tm.     % lambda abstraction
type app  list tm -> tm.              % n-ary application
type all  tm -> (tm -> tm) -> tm.     % forall quantifier
type c    string -> tm.               % constants
\end{elpicode}

\noindent
Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
the concrete syntax to abstract, at the meta level, an expression
\elpiIn{e} over a variable \elpiIn{x}
is \elpiIn{x\ e}, and square brackets denote a list of
terms separated by comma. As an example we show the encoding of the Coq term
``$\forall y:\mathrm{t}.\mathrm{nfact}~ y~ 3$'':

\begin{elpicode}
all (c"t") y\ app[c"nfact", y, c"3"]
\end{elpicode}

\noindent
We now illustrate the encoding of the three instances above as higher order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises;
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

\begin{elpicode}
finite (app[c"fin", N]).                             % r1
decision (app [c"nfact", N, NF]).                    % r2
decision (all A x\ app[P, x]) :- finite A,           % r3
  pi w\ decision (app[P, w]).
\end{elpicode}

\noindent
Unfortunately this direct translation of rule (r3) uses the
predicate \coqIn{P} essentially as a first order term for the meta
language (its type is \elpiIn{tm}).
If we try to backchain the rule (r3) on the encoding of the goal (g) above:

\begin{elpicode}
decision (all (app[c"fin", c"7"]) y\                 % g
  app[c"nfact", y, c"3"]).
\end{elpicode}

\noindent
we fail because of this ``higher order'' unification problem (for DTT)
is phrased as a first order unification problem in the meta language:
the two lists of terms have different lengths!

\begin{elpicode}
app[c"nfact", y, c"3"] = app[P, y]                   % p
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule (r3) as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm A P, finite A,   % r3a
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
in its scope, the unification problem (pa) admits one solution:

\begin{elpicode}
app[c"nfact", y, c"3"] = Pm y                       % pa
Pm = x\ app[c"nfact", x, c"3"]       % assignment for Pm
A = app[c"fin", c"7"]                % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule (r3a) with the goal Elpi runs
the premise \elpiIn{link Pm A P} that is in charge of bringing the
assignment for \elpiIn{Pm} (that has type \elpiIn{tm -> tm})
back to the domain of Coq terms (the type \elpiIn{tm}):

@@

\noindent
This simple example is sufficient to show that the encoding we seek
is not trivial, since the solution for \elpiIn{P} above generates a
(Coq) $\beta$-redex in the second premise (the predicate
under the \elpiIn{pi w\ }\hspace{-0.8em}):

\begin{elpicode}
decision (app[lam A (a\ app[c"nfact", a, c"3"]), w])
\end{elpicode}

\noindent
In turn this redex prevents the rule (r2) to backchain properly since
the following unification problem has no solution:

@@

\noindent
The root cause of the problems we face is that the unification procedure
\Ue of the meta language is not aware of the equational theory of the object
logic \U, even if both theories include $\eta\beta$-conversion and admit most general
unifiers for problems in the pattern fragment~\cite{miller92jsc}.

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi~\ref{sec:encodings}, then we identify a minimal language \Fo
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. TODO citare Teyjus.

\section{Problem statement and alternative encodings} %%%%%%%%%%%%%%%%%%%%%%
\label{sec:encodings}

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to definition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher order problems
restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
We call this unification procedure \U{}.

The equational theory of the meta language Elpi that we want to use to
implement a form of proof automation is strikingly similar, since it
it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher order problems in
$\mathcal{L}_\lambda$.

In spite of the similarity the link between \Ue{} and \U{} is not trivial,
since the abstraction and application term constructors
the two unification procedure deal with are different. For example

\begin{tabular}{lcl}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[c"f", x]} & \U{} & \elpiIn{c"f"}\\
\elpiIn{lam A x\ app[c"f", x]} & \nUe{} & \elpiIn{c"f"} \\
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \U{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}

\noindent
One could ignore this similarity, and ``just'' describe the object language
unification procedure in the meta language, that is crafting a \elpiIn{unif}
predicate to be used as follows in rule (r3):

\begin{elpicode}
decision X :- unif X (all A x\ app[P, x]), finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
This choice would underuse the logic programming engine provided by
the metalanguage since by removing any datum from the head of rules
indexing degenerates. Moreover the unification procedure built in the
meta language is likely to be faster than one implemented in it,
especially if the meta language is interpreted as Elpi is.

To state precisely the problem we solve we need a \Fo{} representation
of DTT terms, a \Ho{} one. Then for any problem $s_1 \U{} s_2$ for
$s_i$ in \Fo{}, we get a  $t_1 \Ue{} t_2$ for $s_i$ in \Fo{} and
a $L = \bigwedge_i \mathrm{link}~ s_i~ t_i $ such that:

\begin{align*}
s_1 \U{} s_2 & \Leftrightarrow  t_1 \Ue{} t_2 \land L \\
s_1 \nU{} s_2 & \Rightarrow t_1 \nUe{} t_2 \lor \lnot L
\end{align*}

where $\mathrm{comp}~t_i = (s_i, \mathrm{link}~ s_i~ t_i)$ and 
link is a predicate allowed to use \U{} on ground terms (hence equal, not unif)

These properties allow us to simulate a unification based
backward search on DTT by using \Ue{}. In particular
any unification that would work in DTT will work,
and whenever a unification in DTT would have failed
either the corresponding \Ue{} or the link fails.

This grants that the trace of the logic program
performing the search not only gives the same result, but
also takes the same paths, that is it fails as early as possible.

\subsection{Alternative encodings and related work}

@@
